A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-29 latex/main.tex

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; 
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(Table~\ref{tab:method_concordance}).
Of the Type I events, 13 are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

@@

@@

%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination,
as measured by the number of resulting mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination,
the stronger the evidence for recombination.
We calculated the number of averted mutations for
all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages})
and those studied by Jackson et al. \cite{Jackson2021}
(Table \ref{tab:jackson}, Document~\ref{sec:jackson_recombs}).

@@

\subsubsection*{Quality control of recombination events}
Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
Fortunately, artefactual recombinants caused by this effect can be identified
because they affect multiple adjacent or near-adjacent bases simultaneously.

@@

>>>>

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; 
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}).
Of the Type I events, 13 are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

@@

\subsection*{Quality control of recombination events}
Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues} (MNSs)
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure 
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues. 
Figure~\ref{fig:recombinant_qc} plots the net number of supporting
loci for the left and right parents for all 855 recombination
events. The 501 recombinants with fewer than 4 net loci supporting 
each parent are highly enriched for the AmpliSeq V1 sequencing
protocol [TODO concrete numbers?], 
and 71 (14\%) are associated with 6-base deletion in the 
Delta lineage, known to be affected 
by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and 
the Jackson et al recombinants (Document\ref{sec:jackson_recombs}).

To assist in the identification of such issues
we developed a SnpIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci.
Each copying pattern shows the positions
where the allelic state in a recombinant
(middle row) matches that of the left (upper row, coloured green) 
or right parent (lower row, coloured blue),
or where the recombination event requires a de-novo mutation
(gold, with mutational change below).
See Figures~\ref{fig:averted_mutations}D and
\ref{fig:pango_x_copying_patterns} for copying patterns of the 
Pango X lineages, and 
Figure~\ref{fig:copying_pattern_examples} for copying patterns
highlighting specific types of QC problem (and more details on the 
visualisation).

@@

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between 
the two solutions---the number of 
mutations ``averted'' by recombination---allows us to quantify
the plausibility.

Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination,
as measured by the number of resulting mutations 
The more mutations are ``averted'' by allowing recombination,
the stronger the evidence for recombination.
We calculated the number of averted mutations for
all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages})
and those studied by Jackson et al. \cite{Jackson2021}
(Table \ref{tab:jackson}, Document~\ref{sec:jackson_recombs}).

@@

\subsubsection*{Quality control of recombination events}
[TODO rewrite this to account for new Results section]

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
Fortunately, artefactual recombinants caused by this effect can be identified
because they affect multiple adjacent or near-adjacent bases simultaneously.

@@

[TODO explain why we restrict to the Type I events]
